---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# geometr <a href='https://ehrmanns.github.io/geometr/'><img src='man/figures/logo.svg' align="right" height="200" /></a>

[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/geometr)](https://cran.r-project.org/package=geometr)
[![Travis-CI Build
Status](https://travis-ci.org/EhrmannS/geometr.svg?branch=master)](https://travis-ci.org/EhrmannS/geometr)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/EhrmannS/geometr?branch=master&svg=true)](https://ci.appveyor.com/project/EhrmannS/geometr)
[![Coverage
Status](https://img.shields.io/codecov/c/github/EhrmannS/geometr/master.svg)](https://codecov.io/github/EhrmannS/geometr?branch=master)
[![](http://cranlogs.r-pkg.org/badges/grand-total/geometr)](https://cran.r-project.org/package=geometr)
[![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)

## Overview

The `geometr` package provides tools that generate and process easily accessible and tidy geometric shapes (of class `geom`). Moreover, it aims to improve interoperability of spatial and other geometric classes. Spatial
classes are typically a collection of geometric shapes (or their vertices) that are accompanied by various metadata (such as attributes and a coordinate reference system). Most spatial classes are thus conceptually quite similar, yet a common standard lacks for accessing features, vertices or the metadata. `Geometr` fills this gap by
providing tools that

  - produce an identical output for the same metadata of different classes (via so-called getters) and
  - use an identical input to write to various classes that originally require different input (via so-called setters).

## Installation

1) Install the official version from CRAN:

```{r, eval=FALSE}
install.packages("geometr")
```

or the latest development version from github:

```{r, eval=FALSE}
devtools::install_github("EhrmannS/geometr")
```

2) The [vignette](https://ehrmanns.github.io/geometr/articles/geometr.html) gives an in depth introduction, explains the take on interoperability and discusses the spatial class `geom` that comes with `geometr`.

3) Have fun being a [geometer](https://en.wikipedia.org/wiki/List_of_geometers)\!

## Examples

Create a `geom`

```{r}
library(geometr)

# ... from other classes
library(sf)
nc_sf <- st_read(system.file("shape/nc.shp", package="sf"), quiet = TRUE)
nc_geom <- gc_geom(input = nc_sf)

# ... or by hand.
library(tibble)
coords <- tibble(x = c(40, 70, 70, 50),
                 y = c(40, 40, 60, 70))
window <- tibble(x = c(0, 80),
                 y = c(0, 80))
aGeom <- gs_polygon(anchor = coords, window = window)

# The "tiny map" shows where the vertices are concentrated.
nc_geom
```

Metadata of different classes can be extracted in interoperable quality (i.e. the same metadata in different objects/classes have the same name and the same arrangement).

```{r}
getFeatures(x = nc_sf)
getFeatures(x = nc_geom)
```

`geometr` only knows the feature types `point`, `line`, `polygon` and `grid` (a systematic lattice of points). 
In contrast to the simple features standard, there are no  *MULTI\** features. 
The way simple features have been implemented in R means that the same information can be stored in several different ways, which are only interoperable after a range of tests and corrections. 
For example, a group of polygons can make up a `MULTIPOLYGON` with attributes that are valid for the overall group only.
Likewise, the polygons could be stored at the level of individual \"closed paths\" as `POLYGON`, with specific attributes per polygon.
Both sets of attributes can only exists either as duplicates for all group specific attributes in a `POLYGON`, or even more complicated nested attribute tables at the `MULTIPOLYGON` level.

The backbone of a `geom` are three attribute tables, one for points, features and groups of features, the latter two of which can be provided with ancillary information.
Each feature is stored as a single unit, all of which are related to other features by a group ID, which relates the features to attributes for an overall group.
Eventually this results in a tidier data-structure with easier access than `Spatial*` of `sf` objects and with higher versatility.

```{r}
# when using the group = TRUE argument, the attributes of MULTI*-feature are
# stored in the group attribute table of a geom
nc_geom <- gc_geom(input = nc_sf, group = TRUE)
currituck <- getFeatures(x = nc_geom, gid == 4)

getFeatures(x = currituck)
getGroups(x = currituck)

# and new attributes can be set easily, 
newTable <- data.frame(fid = c(1:108), 
                       attrib = rnorm(108))
(nc_geom <- setFeatures(x = nc_geom, table = newTable))
```

`geometr` comes with the `visualise` function, which makes nice-looking diagnostic spatial plots, that show explicit values whenever possible.
For example, it does not create artificial bins for the values scale, but shows the explicit range of values.
Moreover, you can easily set plot titles without much effort.

```{r, fig.height=3}
visualise(`North Carolina` = nc_geom)
```

By default, `visualise` uses the feature ID as fillcolour. 
You can use quick options to modify which aspect given object should be shown in the plot, for example to scale the `fillcol`our to the attribute `NWBIR74`.

```{r, fig.height=3}
myAttr <- "NWBIR74"
visualise(!!paste0("NC - ", myAttr) := nc_geom, fillcol = !!myAttr)
```

Each `geom` has the slot `@window`, which contains a reference window. This reference window can be used or modified in many functions of `geometr`

```{r, fig.height=3.5}
visualise(`Currituck` = currituck, window = getExtent(currituck))
```

Finally, cast a `geom` to another type simply by providing it in `anchor` of the respective type

```{r, fig.height=3.5}
library(magrittr)
boundPoints <- gs_point(anchor = currituck) %>% 
   setWindow(to = getExtent(.))
visualise(`Currituck - boundary vertices`= boundPoints, linecol = fid)
```
