---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# geometr <a href='https://ehrmanns.github.io/geometr/'><img src='man/img/logo.svg' align="right" height="200" /></a>

[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/geometr)](https://cran.r-project.org/package=geometr)
[![Travis-CI Build
Status](https://travis-ci.org/EhrmannS/geometr.svg?branch=master)](https://travis-ci.org/EhrmannS/geometr)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/EhrmannS/geometr?branch=master&svg=true)](https://ci.appveyor.com/project/EhrmannS/geometr)
[![Coverage
Status](https://img.shields.io/codecov/c/github/EhrmannS/geometr/master.svg)](https://codecov.io/github/EhrmannS/geometr?branch=master)
[![](http://cranlogs.r-pkg.org/badges/grand-total/geometr)](https://cran.rstudio.com/web/packages/geometr/index.html)
[![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)

## Overview

The `geometr` package provides tools that generate and process easily accessible and tidy geometric shapes (of class `geom`). Moreover, it aims to improve interoperability of spatial and other geometric classes. Spatial
classes are typically a collection of geometric shapes (or their vertices) that are accompanied by various metadata (such as attributes and a coordinate reference system). Most spatial classes are thus conceptually quite similar, yet a common standard lacks for accessing features, vertices or the metadata. `Geometr` fills this gap by
providing tools

  - that produce an identical output for the same metadata of different classes (via so-called getters) and
  - that use an identical input to write to various classes that originally require different input (via so-called setters).

## Installation

1)  Install the development version from github and load it:

```{r, eval=FALSE}
devtools::install_github("EhrmannS/geometr")
```

2)  The [vignette](https://ehrmanns.github.io/geometr/articles/geometr.html) gives a more in depth introduction, explains `geometr`s take on interoperability and discusses the spatial class `geom` that comes with `geometr`.

3)  Have fun being a [geometer](https://en.wikipedia.org/wiki/List_of_geometers)\!

## Examples

Create a `geom`

```{r}
library(geometr)

# ... from other classes
library(sf)
nc_sf <- st_read(system.file("shape/nc.shp", package="sf"), quiet = TRUE)
nc_geom <- gc_geom(input = nc_sf)

# ... or by hand.
library(tibble)
coords <- tibble(x = c(40, 70, 70, 50),
                 y = c(40, 40, 60, 70))
window <- tibble(x = c(0, 80),
                 y = c(0, 80))
aGeom <- gs_polygon(anchor = coords, window = window)

# The "tiny map" shows where the vertices are concentrated.
nc_geom
```

Metadata of different classes can be extracted in interoperable quality (i.e. the same metadata in the same arrangement).

```{r}
getTable(x = nc_sf)
getTable(x = nc_geom, slot = "feature")
```

A `geom` has three attribute tables, one for vertices, one for features and one for groups of features, all of which can be provided with ancillary information ([details on data structure of a `geom`](https://ehrmanns.github.io/geometr/articles/geometr.html#the-class-geom)).

```{r}
getTable(x = nc_geom, slot = "point")
getTable(x = nc_geom, slot = "group")
```

Groups of features are called *multi\** features in other packages. By lumping several closed geometric shapes into one multi\* feature, the separate geometric shapes can’t be attributed with ancillary information any more. In a `geom`, multi\* features are separated into distinct (simpler) features, while the attributes of multi\* features can be captured by the *group attribute table*.

```{r}
nc_geom2 <- gc_geom(input = nc_sf, group = TRUE)
currituck <- getSubset(x = nc_geom2, gid == 4)
getTable(x = currituck, slot = "feature")
getTable(x = currituck, slot = "group")
```

Visualise a `geom`

```{r}
visualise(`North Carolina` = nc_geom)
visualise(`NC - NWBIR74` = nc_geom, fillcol = NWBIR74)
```

A `geom` has the slot `@window`, which contains a reference window, so to speak. This reference window can be used or modified in many functions of `geometr`

```{r}
visualise(`Currituck` = currituck, window = getExtent(currituck))
```

Finally, cast a `geom` to another type simply by providing it in `anchor` of the respective type

```{r}
library(magrittr)
boundPoints <- gs_point(anchor = currituck) %>% 
   setWindow(to = getExtent(.))
visualise(`Currituck - boundary vertices`= boundPoints)
```
