---
title: "Benchmarks"
author: "Steffen Ehrmann"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_caption: true
    theme: spacelab
    highlight: pygments
vignette: >
  %\VignetteIndexEntry{benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This script compares carries out benchmarks for diferent functions.

```{r}
# packages for testing
library(bench)
library(testthat)
library(checkmate)
library(geometr)
library(raster)
```

## Objects to test

First I create a couple of objects that will be tested.

```{r, cache=TRUE}
largeRaster <- getData("worldclim", var = "tmin", res = 10)$tmin1
smallRaster <- gtRasters$categorical

largeGridGeom <- gc_geom(input = largeRaster)
smallGridGeom <- gtGeoms$grid$categorical
```

The objects have the following sizes

```{r}
getSize <- function(x){
  temp <- object.size(x)
  print(temp, units = "KB")
}

getSize(x = largeRaster)
getSize(x = largeGridGeom)

getSize(x = smallRaster)
getSize(x = smallGridGeom)
```

`largeGridGeom` is the largest file, as it is recently loaded into memory entirely.
However, `smallGridGeom` is smaller than `smallRaster`, so probably `largeGridGeom` is also smaller than `largeRaster`, once that is loaded into memory for plotting.

# geom vs raster


## getters

```{r}
mark(getCRS(x = largeGridGeom), getCRS(x = largeRaster), check = FALSE)

mark(getExtent(x = largeGridGeom), getExtent(x = largeRaster), check = FALSE)

mark(getFeatures(x = largeGridGeom), getFeatures(x = largeRaster), check = FALSE)

mark(getGroups(x = largeGridGeom), getGroups(x = largeRaster), check = FALSE)

mark(getHistory(x = largeGridGeom), getHistory(x = largeRaster), check = FALSE)

mark(getPoints(x = largeGridGeom), getPoints(x = largeRaster), check = FALSE)

mark(getRes(x = largeGridGeom), getRes(x = largeRaster), check = FALSE)

mark(getType(x = largeGridGeom), getType(x = largeRaster), check = FALSE)

mark(getWindow(x = largeGridGeom), getWindow(x = largeRaster), check = FALSE)
```

## visualise

```{r, fig.show='hide', warning=FALSE}
mark(visualise(smallGridGeom), visualise(smallRaster), check = FALSE, iterations = 10)
mark(visualise(smallRaster), plot(smallRaster), check = FALSE, iterations = 10)
```

```{r, fig.show='hide', warning=FALSE, cache=TRUE}
mark(visualise(largeGridGeom), visualise(largeRaster), check = FALSE, iterations = 10)
mark(visualise(largeRaster), plot(largeRaster), check = FALSE, iterations = 10)
```
