---
title: "geometr - An overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Interoperability

R is an open source environment which means that no single package or class will ever be the sole source of a particular datastructure and this is also the case for spatial or geometric data.

Interoperable software is software that can easily exchange information with other software, which is typically achieved by providing the output of functionally similar operations in a common arrangement or format.
`Geometr` improves interoperability in R for spatial classes, by providing functions that extract and provide metadata in an identical format from a potentially wide range of classes.

Think, for example about the extent of a spatial object.
The extent is characterised by a minimum and maximum value in all dimensions into which the object spans.
There are, however, several forms in which this extent can be reported, such as a vector of four numbers or a 2x2 table.
When it comes to the structure of the reported information, several questions arise:

* Which value is reported in which position?
* Does the reported object have names and if so, what do they mean?
* How do names/positions relate to names/positions of objects reported by other software?

Take the following examples, which are all considered standard in R to date:

```{r}
library(sf)
nc <- st_read(system.file("shape/nc.shp", package="sf"))
st_bbox(nc)

library(sp)
nc_sp <- as_Spatial(nc)
bbox(nc_sp)

library(raster)
ras <- raster(system.file("external/test.grd", package="raster"))
extent(ras)
```

For a human user the structure of those information might not matter.
The capacity of humans to recognise information relative to their context needs to be programmed into the software that is supposed to gather those information.
Think of a third function that is supposed to take some spatial input from the user and derive information from it, without knowing in advance what spatial class that will be, and hence does not know how the output of the respective class should be interpreted.
The need for such a function is derived from the different flavours/functions/pipelines that are common in an open source software, so that all classes, which may be preferable in particular usecases, can be used equally.
So far, this interoperability is not given in R and `geometr` provides it, as the following example shows

```{r}
library(geometr)
myInput <- nc
getExtent(x = myInput)

myInput <- nc_sp
getExtent(x = myInput)

myInput <- ras
getExtent(x = myInput)
```

This ensures that a custom function that is meant to process several spatial classes, and which must thus be agnostic to the class specific peculiarities, requires merely one very simple row of code and not a lengthy if-condition that tests classes and implements different extraction procedures.
Moreover, the output of the tools provided by `geometr` is tidy, i.e. it provides variables in columns and observations in rows.


## The class `geom`