---
title: "getters_setters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getters_setters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(geometr)
library(tibble)
```

## Getters and Setters

Most properties of a `geom` are accessible via specific functions. They can thus be extracted, modified and re-assigned, which is the basis for various "meta-operations", such as reprojecting or appending additional attributes to features.

```{r}
# # Get properties of a geom
# getVertices(x = gtGeoms$polygon)
# getTable(x = gtGeoms$polygon)
# getWindow(x = gtGeoms$polygon)
# getCRS(x = gtGeoms$polygon) # does not contain any value and is thus only valid in the cartesian coordinate system
# getHistory(x = gtGeoms$polygon)
# 
# # Set properties of a geom ...
# # ... such as feature attributes
# newAttrs <- tibble(fid = c(2, 1), weight = c(10, 5))
# newGeom <- setTable(x = gtGeoms$polygon, table = newAttrs)
# getTable(x = newGeom)
# visualise(geom = newGeom, fillcol = weight)
```

In an attempt to standardise the handling of various different spatial formats, these functions have been developed as generics and they come with a couple of methods for the most-used spatial classes, such as `Raster*`, `sp`, `sf`. Originally this was developed to accommodate the modular nature of `rasterTools` but is now offered as a stand-alone. One can for instance use `getVertices()` to get a standardised output of the coordinate values across all supported spatial classes.

```{r}
# getVertices(gtGeoms$point)
# getVertices(gtSP$SpatialMultiPoints)
# getVertices(gtSF$multipoint)
```
